---
title: "Runs of homozygosity on 943 individuals"
output: github_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,dev = 'jpeg')
suppressMessages(library(knitr))
```


```{r, echo=FALSE}
suppressMessages(library(tidyverse))
suppressMessages(library(gridExtra))
suppressMessages(library(grid))
suppressMessages(library(data.table))
suppressMessages(library(GenomicRanges))

outlier <- function (x) {
  IQR <- quantile(x)[[4]] - quantile(x)[[2]]
  low <- quantile(x)[[2]] - 1.5*IQR
  high <- quantile(x)[[4]] + 1.5*IQR
  out_val <- x[x < low | x > high]
  return(out_val)
}
```



```{r echo=FALSE}
roh <- fread('../input/roh_concat.txt', showProgress = T)
colnames(roh) <- c("Chromosome", "Start", 'End', 'Length', 'Number_of_markers', 'Quality','sample_id')
roh$sample_id <-gsub(".*/(.*)\\..*", "\\1", roh$sample_id)
roh$sample_id <-gsub(".norm.roh", "", roh$sample_id)
to_include <- read.table('SAMPLES_TO_INCLUDE_210519.txt',sep='\t')

roh <- roh %>% filter(sample_id %in% to_include$V1) %>% filter(Quality > 25 & Number_of_markers >50) 

kable(roh %>% select(Length,Number_of_markers,Quality) %>%
  pivot_longer(cols = everything(), names_to = 'stat',values_to = 'value') %>%
  group_by(stat) %>%
  summarise(min = round(min(value),2),median=round(median(value),2),mean=round(mean(value),2),
            max=round(max(value),2))
)

```

ROHs quality histogram

```{r echo=FALSE, eval=F}
roh  %>% ggplot(aes(x=1,y=Quality)) +
    geom_violin(fill='#48C095', col='#27384A') +
  geom_boxplot(width=0.1) +
     theme_classic() + theme(axis.text.x = element_blank()) +
  xlab('')
```

## Results

1. Cummulative sum

```{r cummulative_sum, echo=FALSE, warning=F}

roh_cummulative <- roh %>% 
  filter(Chromosome != 'chrX' & Chromosome != 'chrY') %>% 
  arrange(Length) %>%
  group_by(sample_id) %>% 
  summarise(Length, LS = cumsum(Length)/1e6)

ranges <- c(1,
            1e4,2e4,5e4,
            1e5,2e5,5e5,
            1e6,2e6,5e6,
            1e7,2e7)

roh_cummulative$Range <- sapply(roh_cummulative$Length, 
                                function(x) max(ranges[x > ranges]))
rc.by.range = roh_cummulative %>% 
              group_by(sample_id, Range) %>%
              summarize(maxLS=max(LS))

rc.by.range.full = data.frame(sample_id=c(), Range=c(), maxLS=c())
css.empty = rep(NA, length(ranges))
names(css.empty) = ranges 
for (sample in unique(rc.by.range$sample_id)) {
  #message(sample)
  t = rc.by.range %>% filter(sample_id==sample)
  css = css.empty
  css[as.character(t$Range)] = t$maxLS
  for (i in c(1:length(css))) {
    if (is.na(css[i])) {
      css[i] = ifelse(i==1, 0, css[i-1])
    }
  } 
  rc.by.range.full <- rbind(rc.by.range.full, data.frame(sample_id=sample, 
                                                         Range = names(css),
                                                         maxLS = css))
}
       

plot_data <- rc.by.range.full %>% 
              group_by(Range) %>%
              summarise(quantiles = scales::percent(c(0.1, 0.25, 0.5, 0.75, 0.9)),
                        quantile_length = quantile(maxLS, c(0.1, 0.25, 0.5, 0.75, 0.9)))
            

q25 <- plot_data %>% filter(quantiles=='25%')
q75 <- plot_data %>% filter(quantiles=='75%')
q10 <- plot_data %>% filter(quantiles=='10%')
q90 <- plot_data %>% filter(quantiles=='90%')
med <- plot_data %>% filter(quantiles=='50%')


ggplot(med, aes(x=as.numeric(Range), y=quantile_length, col='#48C095', linetype='solid')) + geom_line() +
  geom_line(aes(x=as.numeric(q25$Range),y=q25$quantile_length, col='#27384A',linetype='dashed')) +
  geom_line(aes(x=as.numeric(q75$Range),y=q75$quantile_length, col='#27384A',linetype='dashed')) +
  geom_line(aes(x=as.numeric(q10$Range),y=q10$quantile_length, col='#27384A',linetype='dotted')) +
  geom_line(aes(x=as.numeric(q90$Range),y=q90$quantile_length, col='#27384A',linetype='dotted')) +
  scale_x_log10(breaks=c(0.01, 0.1, 0.5, 1, 5, 10)*1e6,
                labels=c('<10kb', '<100kb', '<500kb', '<1Mb', '<5Mb', '<10Mb'), 
                limits=c(1e4, 2e7)) + theme(legend.position = 'top') +
  ylab('Cummulative size of ROH (Mb)') +
  xlab('ROH length (Mb)') +   scale_color_identity('none') +
  scale_linetype_identity(name = "",
                       breaks = c("solid", "dashed", "dotted"),
                       labels = c("Median", "Quantiles 25 & 75", "Quantiles 10 & 90"),
                       guide = "legend")+
  theme(legend.position = c(0.87, 0.25), legend.background = element_rect(fill = "transparent", color = "black"))



```

2. Median SROH

```{r Median_SROH,echo=FALSE}
roh_total <- roh %>%
  filter(Chromosome != 'chrX' & Chromosome != 'chrY') %>% arrange(Length)
ranges <- c(0.5,1,2,4,8,16)

roh_total$Range <- NA
roh_total$Length <- roh_total$Length/1e+06
roh_total <- roh_total %>% filter(Length >= 0.5)

for (i in 1:length(ranges)) {
  if (ranges[i] != max(ranges)) {
      roh_total$Range <- ifelse(roh_total$Length > ranges[i] & 
                                        roh_total$Length < ranges[i+1], paste(
                                          ranges[i],'-',ranges[i+1],'Mb',sep=''
                                        ), roh_total$Range) 
  } else {
    roh_total$Range <- ifelse(roh_total$Length >= ranges[i], paste('>',ranges[i],'Mb',sep='' ), roh_total$Range) 
      
  }
}

roh_total$Range <- factor(roh_total$Range)
roh_total$Range <- ordered(roh_total$Range, levels = c("0.5-1Mb", "1-2Mb", "2-4Mb", "4-8Mb",
                                                         "8-16Mb", ">16Mb" ))

roh_total <- roh_total %>% group_by(Range,sample_id) %>%
  summarise(Length = sum(Length))

plot_data <- roh_total %>% group_by(Range) %>%
  summarise(quantiles = scales::percent(c(0.25, 0.5, 0.75)),
            quantile_length = quantile(Length, c(0.25, 0.5, 0.75))
            )

q1 <- plot_data %>% filter(quantiles=='25%')
q3 <- plot_data %>% filter(quantiles=='75%')
med <- plot_data %>% filter(quantiles=='50%')


ggplot(med,aes(x=as.numeric(Range),y=quantile_length)) +
  geom_line(col='#48C095') +
  geom_line(aes(x=as.numeric(q1$Range),y=q1$quantile_length),col='#27384A',linetype='dashed') +
  geom_line(aes(x=as.numeric(q3$Range),y=q3$quantile_length),col='#27384A',linetype='dashed') +
  theme_classic() +
  scale_x_continuous(breaks=seq(1,length(med$Range),1),
                     labels=med$Range[seq(1,length(med$Range),1)]) + 
  ylab('Median SROH per individual (Mb)') +
  xlab('ROH length category (Mb)') +
  theme(legend.position = "none")
```


2. Coverage heatmap

```{r genome_coverage_heatmap, echo=FALSE}
library(GenomicRanges)
c = coverage(GRanges(seqnames = roh$Chromosome, 
                     ranges=IRanges(roh$Start, roh$End)))

window.size=1e6
m = matrix(nrow = 2*length(names(c)), 
           ncol=sum(runLength(c[[1]]))/window.size)

for (chr in 1:22) { #seqnames(gr)) {
  max = sum(runLength(c[[chr]]))
  sampling.points = IRanges(start=c(1:(max/1e6))*1e6, width = 1)  
  a = aggregate(c[[chr]], sampling.points, mean)
  m[(chr*2)-1,1:length(a)] = a
}

image(t(m[dim(m)[1]:1,]), xaxt='n', yaxt='n')

```


### All results below are ROHs filter for autosomes with Quality > 25%, Number of markers > 50 and ROH length > 2 Mb

3. Average sum of ROHs per genome

```{r total_roh, echo=FALSE}
roh <- roh  %>%
  filter(Length <= 5e+06 & Length >=2e+06 | Length > 10e+06) %>% 
  filter(Chromosome != 'chrX' & Chromosome != 'chrY') %>% arrange(Length)

roh$Range <- NA
roh$Range <- ifelse(roh$Length <= 5e+06 & roh$Length >=2e+06, '2Mb-5Mb',roh$Range)
roh$Range <- ifelse(roh$Length > 10e+06, '>10Mb',roh$Range)

roh$Range <- factor(roh$Range)
roh$Range <- ordered(roh$Range, levels = c('2Mb-5Mb','>10Mb'))

group.colors <- c('>10Mb' = '#48C095', '2Mb-5Mb' = '#27384A')


#roh$Chromosome <- gsub('chr','',roh$Chromosome)
#roh$Chromosome <- as.numeric(roh$Chromosome)

roh$Chromosome <- factor(roh$Chromosome)
roh$Chromosome <- ordered(roh$Chromosome, levels = unique(roh$Chromosome))

roh %>% group_by(sample_id) %>%
  summarise(SROH_Mb = sum(Length)/1e+06) %>%
  ggplot(aes(y=1,x=SROH_Mb)) +
  geom_violin(fill='#48C095', col='#27384A') +
  geom_boxplot(width=0.1) +
  theme_classic() +
  xlab('SROH (Mb)') +
  ylab('') +
  theme(axis.text.y = element_blank())

kable(summary(
  roh %>% group_by(sample_id) %>%
  summarise(mean_SROH_Mb = sum(Length)/1e+06) %>%
    select(mean_SROH_Mb)
  )
  
)

kable(
  roh %>% group_by(Range) %>%
    summarise(mean_length = mean(Length)/1e+06)
)
```

4. Sum of ROHs per chromosome and individual

```{r, echo=F}
roh$Chromosome <- gsub('chr','',roh$Chromosome)
roh$Chromosome <- as.numeric(roh$Chromosome)

roh$Chromosome <- factor(roh$Chromosome)
roh$Chromosome <- ordered(roh$Chromosome, levels = 1:22)

roh %>% group_by(sample_id,Chromosome,Range) %>%
  arrange(desc(Chromosome)) %>%
  summarise(SROH_Mb = sum(Length)/1e+06) %>%
  ggplot(aes(x=Chromosome,y=SROH_Mb,col=Range)) +
  geom_jitter() +
  theme_classic() +
  ylab('SROH (Mb)') +
  xlab('Chromosome') +
  scale_color_manual(values=group.colors)

kable(summary(
  roh %>% group_by(sample_id,Chromosome) %>%
  summarise(mean_SROH_Mb = sum(Length)/1e+06) %>%
    select(mean_SROH_Mb)
  )
  
)
```


5. Sum of ROH length per range

```{r SROH, echo=FALSE}
sroh <- roh %>% filter(Chromosome != 'chrX' & Chromosome != 'chrY') %>%
  filter(Length >= 2e+06 & Length <= 5e+06 | Length >= 10e+06 & Number_of_markers >= 50)
  
sroh_sum <- sroh %>% group_by(Range,sample_id) %>% summarise(SROH = sum(Length)/1e+06)

sroh_sum %>% ggplot(aes(y=Range,x=SROH)) +
 geom_violin(fill='#48C095', col='#27384A') +
 geom_boxplot(width=0.1) +
 theme_classic() +
 ylab('Range') +
 xlab('SROH (Mb)')

kable(
  summary(sroh_sum %>% 
            pivot_wider(names_from = Range, values_from = SROH) %>%
            select(-sample_id) %>% na.omit()
          )
)


```


6. Relationship between number of ROHs and total length of genome covered by them

```{r SROH_corr, echo=FALSE} 
croh <- roh %>% filter(Chromosome != 'chrX' & Chromosome != 'chrY' & Number_of_markers >= 50) 

croh_sum <- croh %>% group_by(sample_id) %>% 
  summarise(n = n(), sum = sum(Length)/1e+06)

croh_sum %>% ggplot(aes(y=n,x=sum)) +
  geom_point(col='#48C095') +
  theme_classic() +
  ylab('Number of ROH') +
  xlab('Total sum of ROH > 2Mb (SROH)') +
  ylim(c(0,40)) +
  xlim(c(0,300))
```


<!-- 6. ROH genome coverage  -->

<!-- ```{r genome_coverage, echo=FALSE} -->
<!-- chr <- 'chr1' -->
<!-- coverage <- roh %>% filter(Chromosome %in% chr) %>% -->
<!--   arrange(Chromosome,Start) -->
<!-- # coverage$Chromosome <- paste(coverage$Chromosome,sep='') -->
<!-- # coverage$x_from <- paste(coverage$Chromosome,coverage$Start,sep='') -->
<!-- # coverage$x_end <- paste(coverage$Chromosome,coverage$End,sep='') -->



<!-- coverage %>% ggplot() + -->
<!--   geom_segment(aes(x = Start, y = sample_id, xend = End,  -->
<!--                                  yend = sample_id), col='#48C095') + -->
<!--   theme_classic() + -->
<!--   theme(axis.text.x = element_blank(),axis.text.y = element_blank()) + -->
<!--   xlab('Position on chromosome 1') + -->
<!--   ylab('Individual') -->

<!-- ``` -->


7. % of ROhs per category in sample

```{r roh_count, echo=FALSE}
suppressMessages(
roh_count <- roh  %>% 
 group_by(Range,sample_id,.drop = F) %>% 
 summarise(n=n()) %>% arrange(Range,n)
)

roh_count %>% 
 ggplot(aes(x=reorder(sample_id,-n),y=n,fill=Range)) +
 geom_bar(position="fill", stat="identity",width = 1) +
 scale_fill_manual(values=group.colors) +
 theme_classic() +
 xlab('Individuals') +
 ylab('% of ROHs count') +
 theme(axis.text.x = element_blank())
```

8. ROH length sum

```{r roh_sum, echo=FALSE}
suppressMessages(
roh_sum <- roh  %>% 
 group_by(Range,sample_id,.drop = F) %>% 
 summarise(sum=sum(Length)) %>% arrange(Range,sum)
)

roh_sum %>% 
 ggplot(aes(x=reorder(sample_id,-sum),y=sum,fill=Range)) +
 geom_bar(position="fill", stat="identity",width = 1) +
 scale_fill_manual(values=group.colors) +
 theme_classic() +
 xlab('Individuals') +
 ylab('% of ROHs length sum') +
 theme(axis.text.x = element_blank())
```

<!-- 7. Number of ROHs per sample -->

<!-- ```{r roh_count_per_sample, echo=FALSE}  -->
<!-- suppressMessages( -->
<!-- roh_avg <- roh %>% group_by(sample_id, Range) %>% -->
<!--   summarise(avg_length=mean(Length), n=n()) %>% -->
<!--   arrange(Range,n) -->
<!-- ) -->

<!-- roh_avg %>% -->
<!--   ggplot(aes(x=1:nrow(roh_avg),y=n)) + -->
<!--   geom_point(col='#48C095',alpha=0.5) + -->
<!--   facet_wrap(~Range, nrow = 3,scales = 'free') +  -->
<!--   ylab('Number of ROHs') + -->
<!--   xlab('Individual') + -->
<!--   theme_classic() + -->
<!--   theme(axis.text.x = element_blank()) -->

<!-- ``` -->

<!-- 8. Average ROHs per sample -->

<!-- ```{r average_roh_per_sample, echo=FALSE}  -->
<!-- roh_avg <- roh %>% group_by(sample_id, Range) %>% -->
<!--   summarise(avg_length=mean(Length), n=n()) %>% -->
<!--   arrange(Range,avg_length) -->

<!-- roh_avg %>% -->
<!--   ggplot(aes(x=1:nrow(roh_avg),y=avg_length)) + -->
<!--   geom_point(col='#48C095',alpha=0.5) + -->
<!--   facet_wrap(~Range, nrow = 3,scales = 'free') +  -->
<!--   ylab('Mean ROH length') + -->
<!--   xlab('Individual') + -->
<!--   theme_classic() + -->
<!--   theme(axis.text.x = element_blank()) -->

<!-- ``` -->

<!--
9. Cosanguinity in population
-->
```{r, echo=FALSE, eval=F}
froh <- read.table('../output/roh/froh.output',header=T,sep='\t')

froh %>% ggplot(aes(x=1,y=Froh_genome)) +
    geom_violin(fill='#48C095', col='#27384A') +
  geom_boxplot(width=0.1) +
     theme_classic() + theme(axis.text.x = element_blank()) +
  xlab('')

```