---
title: "Runs of homozygosity on 1076 individuals"
output: github_document
---

## Runs of homozygozity (ROH)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,dev = 'jpeg')
suppressMessages(library(knitr))
```


```{r, echo=FALSE}
suppressMessages(library(tidyverse, warn.conflicts = F))
options(dplyr.summarise.inform = FALSE)
# suppressMessages(library(gridExtra))
# suppressMessages(library(grid))
suppressMessages(library(data.table))
suppressMessages(library(GenomicRanges))
suppressMessages(library(karyoploteR))


outlier <- function (x) {
  IQR <- quantile(x)[[4]] - quantile(x)[[2]]
  low <- quantile(x)[[2]] - 1.5*IQR
  high <- quantile(x)[[4]] + 1.5*IQR
  out_val <- x[x < low | x > high]
  return(out_val)
}
```


```{r, echo=F}
roh <- fread('../input/roh_concat.txt', showProgress = T)
colnames(roh) <- c("Chromosome", "Start", 'End', 'Length', 'Number_of_markers', 'Quality','sample_id')
#roh$sample_id <-gsub(".*/(.*)\\..*", "\\1", roh$sample_id)
#roh$sample_id <-gsub(".norm.roh", "", roh$sample_id)
to_include <- read.table('../../../genom_polaka/SAMPLES_TO_INCLUDE_210716.txt',sep='\t')

roh <- roh %>% filter(sample_id %in% to_include$V1) %>% filter(Chromosome != 'chrX' & Chromosome != 'chrY')
roh$Chromosome <- factor(roh$Chromosome, levels = paste('chr',1:22,sep=''))

```

#### ROHs quality histogram

```{r roh_quality, echo=FALSE, eval=F}
roh %>% ggplot(aes(x=1,y=Quality)) +
    geom_violin(fill='gray', col='#27384A') +
  geom_boxplot(width=0.1) +
     theme_minimal() + theme(axis.text.x = element_blank()) +
  xlab('')

```

```{r}
# Quality filtering: Quality>25 & markers > 50
roh <- roh %>% 
  filter(Quality > 25 & Number_of_markers >50)


```

```{r echo=FALSE, eval=F}
hist(log10(roh$Length), 
     main='Distribution of RoH lengths', 
     xlab='Log10(Length[bp])')
```

#### ROH stats after quality filtering

```{r all.roh.stats, echo=FALSE}

kable(roh %>% select(Length,Number_of_markers,Quality) %>%
  pivot_longer(cols = everything(), names_to = 'stat',values_to = 'value') %>%
  group_by(stat) %>%
  summarise(min = round(min(value),2),
            median=round(median(value),2),
            mean=round(mean(value),2),
            max=round(max(value),2))
)

```


Average ROH size per genome (additionally filtered: >100kb)

```{r stats.per.genome, echo=F}
kable(roh %>% 
        filter(Length>1e5) %>% 
        select(Length,sample_id) %>%
        group_by(sample_id) %>%
        summarise(minLen = round(min(Length),2),
            medianLen=round(median(Length),2),
            meanLen=round(mean(Length),2),
            maxLen=round(max(Length),2),
            count=n(), 
            sumLen = sum(Length)) %>%
  pivot_longer(cols = c(minLen, medianLen, meanLen, maxLen, count, sumLen), names_to = 'stat', values_to = 'value') %>%
  group_by(stat) %>%
  summarise(average = round(mean(value),2))
)
```

## Results

1. Cummulative sum

```{r cummulative_sum, echo=FALSE, warning=F}

roh_cummulative <- roh %>% 
  arrange(Length) %>%
  group_by(sample_id) %>% 
  summarise(Length, LS = cumsum(Length)/1e6)

ranges <- c(1,
            1e4,2e4,5e4,
            1e5,2e5,5e5,
            1e6,2e6,5e6,
            1e7,2e7)

roh_cummulative$Range <- sapply(roh_cummulative$Length, 
                                function(x) max(ranges[x > ranges]))
rc.by.range = roh_cummulative %>% 
              group_by(sample_id, Range) %>%
              summarize(maxLS=max(LS))

rc.by.range.full = data.frame(sample_id=c(), Range=c(), maxLS=c())
css.empty = rep(NA, length(ranges))
names(css.empty) = ranges 
for (sample in unique(rc.by.range$sample_id)) {
  #message(sample)
  t = rc.by.range %>% filter(sample_id==sample)
  css = css.empty
  css[as.character(t$Range)] = t$maxLS
  for (i in c(1:length(css))) {
    if (is.na(css[i])) {
      css[i] = ifelse(i==1, 0, css[i-1])
    }
  } 
  rc.by.range.full <- rbind(rc.by.range.full, data.frame(sample_id=sample, 
                                                         Range = names(css),
                                                         maxLS = css))
}
       

plot_data <- rc.by.range.full %>% 
              group_by(Range) %>%
              summarise(quantiles = scales::percent(c(0.1, 0.25, 0.5, 0.75, 0.9)),
                        quantile_length = quantile(maxLS, c(0.1, 0.25, 0.5, 0.75, 0.9))) %>%
  mutate(Range = as.numeric(as.character(Range)))
            

q25 <- plot_data %>% filter(quantiles=='25%')
q75 <- plot_data %>% filter(quantiles=='75%')
q10 <- plot_data %>% filter(quantiles=='10%')
q90 <- plot_data %>% filter(quantiles=='90%')
med <- plot_data %>% filter(quantiles=='50%')


ggplot(med, aes(x=as.numeric(Range), y=quantile_length, col='gray', linetype='solid')) + geom_line() +
  geom_line(aes(x=as.numeric(q25$Range),y=q25$quantile_length, col='#27384A',linetype='dashed')) +
  geom_line(aes(x=as.numeric(q75$Range),y=q75$quantile_length, col='#27384A',linetype='dashed')) +
  geom_line(aes(x=as.numeric(q10$Range),y=q10$quantile_length, col='#27384A',linetype='dotted')) +
  geom_line(aes(x=as.numeric(q90$Range),y=q90$quantile_length, col='#27384A',linetype='dotted')) +
  scale_x_log10(breaks=c(0.01, 0.02, 0.05, 0.1, 0.2, 0.5, 1, 2, 5, 10)*1e6,
                labels=c('<10kb', '<20kb', '<50kb', '<100kb', '<200kb', '<500kb', '<1Mb', '<2Mb','<5Mb', '<10Mb'),
                limits=c(1e4, 2e7)) + theme(legend.position = 'top') +
  ylab('Cummulative size of ROH (Mb)') +
  xlab('ROH length (Mb)') +   scale_color_identity('none') +
  scale_linetype_identity(name = "",
                       breaks = c("solid", "dashed", "dotted"),
                       labels = c("Median", "Quantiles 25 & 75", "Quantiles 10 & 90"),
                       guide = "legend")+
  theme_minimal() +
  theme(legend.position = c(0.87, 0.25), legend.background = element_rect(fill = "transparent", color = "black"))
  



```

Fig. Relation between size of runs of homozygozity and cumulative length of RoHs in an average individual.

```{r lengths_per_chromosome, echo=FALSE, warning=F}
roh_chrom_lengths <- roh %>% 
  group_by(sample_id, Chromosome) %>% 
  summarise(cnt=n(), 
            meanLen=sum(Length)/1e6)

roh_chrom_lengths$Chromosome <- gsub(roh_chrom_lengths$Chromosome,pattern = 'chr',replacement = '')
roh_chrom_lengths$Chromosome <- factor(roh_chrom_lengths$Chromosome, levels = 1:22)

ggplot(roh_chrom_lengths, aes(x=Chromosome, y=meanLen)) + 
  geom_boxplot(fill='gray') + 
  xlab('Chromosome') + 
  ylab('Mean RoH length (Mb)') +
  theme_minimal()
```

Fig. Distribution of mean length of runs of homozygozity per chromosome

```{r numbers_per_chromosome, echo=FALSE, warning=F}
ggplot(roh_chrom_lengths, aes(x=Chromosome, y=cnt)) + 
  geom_boxplot(fill='gray') + xlab('Chromosome') + ylab('Count of RoH') +
  theme_minimal()
```

Fig. Distribution of counts of runs of homozygozity per chromosome


2. Coverage heatmap

```{r genome_coverage_heatmap, echo=FALSE}
regions <- GRanges(seqnames = roh$Chromosome, 
                     ranges=IRanges(roh$Start, roh$End))

kp <- plotKaryotype(chromosomes = 'autosomal')

kpPlotCoverage(kp,regions)

```

Fig. Chromosomal location of run of homozygozity hotspots - regions frequently covered by runs of homozygozity in this cohort. 

### All results below are ROHs filter for autosomes with Quality > 25%, Number of markers > 50 and ROH length > 2 Mb

2. Average sum of ROHs per genome

```{r total_roh, echo=FALSE}
roh <- roh  %>%
  filter((Length <= 5e+06 & Length >=2e+06) | Length > 10e+06) %>% arrange(Length)

roh$Range <- NA
roh$Range <- ifelse(roh$Length <= 5e+06 & roh$Length >=2e+06, '2Mb-5Mb',roh$Range)
roh$Range <- ifelse(roh$Length > 10e+06, '>10Mb',roh$Range)

roh$Range <- factor(roh$Range)
roh$Range <- ordered(roh$Range, levels = c('2Mb-5Mb','>10Mb'))

group.colors <- c('>10Mb' = 'gray', '2Mb-5Mb' = '#27384A')

roh$Chromosome <- factor(roh$Chromosome)
roh$Chromosome <- ordered(roh$Chromosome, levels = unique(roh$Chromosome))

roh %>% group_by(sample_id) %>%
  summarise(SROH_Mb = sum(Length)/1e+06) %>%
  ggplot(aes(y=1,x=SROH_Mb)) +
  geom_violin(fill='gray', col='#27384A') +
  geom_boxplot(width=0.1) +
  theme_classic() +
  xlab('SROH (Mb)') +
  ylab('') +
  theme(axis.text.y = element_blank())

kable(summary(
  roh %>% group_by(sample_id) %>%
  summarise(mean_SROH_Mb = sum(Length)/1e+06) %>%
    select(mean_SROH_Mb)
  )

)

kable(
  roh %>% group_by(Range) %>%
    summarise(mean_length = mean(Length)/1e+06)
)
```




3. Sum of ROH length per range

```{r SROH, echo=FALSE}
sroh_sum <- roh %>% 
  filter(((Length >= 2e+06 & Length <= 5e+06) | Length >= 10e+06)) %>% 
  group_by(Range,sample_id) %>% summarise(SROH = sum(Length)/1e+06)

sroh_sum %>% ggplot(aes(y=Range,x=SROH)) +
 geom_violin(fill='gray', col='#27384A') +
 geom_boxplot(width=0.1) +
 theme_classic() +
 ylab('Range') +
 xlab('SROH (Mb)')

kable(
  summary(sroh_sum %>% 
            pivot_wider(names_from = Range, values_from = SROH) %>%
            select(-sample_id) %>% na.omit()
          )
)


```


4. Relationship between number of ROHs and total length of genome covered by them

```{r SROH_corr, echo=FALSE} 

croh_sum <- roh %>% group_by(sample_id) %>% 
  summarise(n = n(), sum = sum(Length)/1e+06)

croh_sum %>% ggplot(aes(y=n,x=sum)) +
  geom_point(col='gray') +
  theme_classic() +
  ylab('Number of ROH') +
  xlab('Total sum of ROH > 2Mb (SROH)') +
  ylim(c(0,40)) +
  xlim(c(0,300))
```
