---
title: "Runs of homozygosity on 943 individuals"
output: github_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressMessages(library(knitr))
```


```{r, echo=FALSE}
suppressMessages(library(tidyverse))
suppressMessages(library(gridExtra))
suppressMessages(library(grid))
suppressMessages(library(data.table))
suppressMessages(library(GenomicRanges))

outlier <- function (x) {
  IQR <- quantile(x)[[4]] - quantile(x)[[2]]
  low <- quantile(x)[[2]] - 1.5*IQR
  high <- quantile(x)[[4]] + 1.5*IQR
  out_val <- x[x < low | x > high]
  return(out_val)
}
```



```{r echo=FALSE}
roh <- fread('../input/roh_concat.txt', showProgress = T)
colnames(roh) <- c("Chromosome", "Start", 'End', 'Length', 'Number_of_markers', 'Quality','sample_id')
roh$sample_id <-gsub(".*/(.*)\\..*", "\\1", roh$sample_id)
roh$sample_id <-gsub(".norm.roh", "", roh$sample_id)
to_include <- read.table('../../../genom_polaka/SAMPLES_TO_INCLUDE_210519.txt',sep='\t')

roh <- roh %>% filter(sample_id %in% to_include$V1)



kable(roh %>% select(Length,Number_of_markers,Quality) %>%
  pivot_longer(cols = everything(), names_to = 'stat',values_to = 'value') %>%
  group_by(stat) %>%
  summarise(min = round(min(value),2),median=round(median(value),2),mean=round(mean(value),2),
            max=round(max(value),2))
)

```

ROHs quality histogram

```{r echo=FALSE}
roh  %>% ggplot(aes(x=1,y=Quality)) +
    geom_violin(fill='#48C095', col='#27384A') +
  geom_boxplot(width=0.1) +
     theme_classic() + theme(axis.text.x = element_blank()) +
  xlab('')
```

## Results

### All results below are ROHs filter for autosomes with Quality > 25% 

1. Average sum of ROHs per genome

```{r total_roh, echo=FALSE}
roh <- roh  %>% filter(Quality > 25 ) %>%
  filter(Chromosome != 'chrX' & Chromosome != 'chrY')
roh$Range <- NA
roh$Range <- ifelse(roh$Length > 1e+06, '>1Mb',roh$Range)
roh$Range <- ifelse(roh$Length <= 1e+06 & roh$Length >=25e+03, '25kb-1Mb',roh$Range)
roh$Range <- ifelse(roh$Length < 25e+03, '<25kb',roh$Range)

roh$Range <- factor(roh$Range)
roh$Range <- ordered(roh$Range, levels = c('<25Kb','25kb-1Mb','>1Mb'))

group.colors <- c('>1Mb' = '#48C095', '25kb-1Mb' = '#27384A', '<25kb' = '#B6B6B6')


#roh$Chromosome <- gsub('chr','',roh$Chromosome)
#roh$Chromosome <- as.numeric(roh$Chromosome)

roh$Chromosome <- factor(roh$Chromosome)
roh$Chromosome <- ordered(roh$Chromosome, levels = unique(roh$Chromosome))

roh %>% group_by(sample_id) %>%
  summarise(SROH_Mb = sum(Length)/1e+06) %>%
  ggplot(aes(y=1,x=SROH_Mb)) +
  geom_violin(fill='#48C095', col='#27384A') +
  geom_boxplot(width=0.1) +
  theme_classic() +
  xlab('SROH (Mb)') +
  ylab('') +
  theme(axis.text.y = element_blank())

kable(summary(
  roh %>% group_by(sample_id) %>%
  summarise(mean_SROH_Mb = sum(Length)/1e+06) %>%
    select(mean_SROH_Mb)
  )
  
)

kable(
  roh %>% group_by(Range) %>%
    summarise(mean_length = mean(Length)/1e+06)
)
```

2. Sum of ROHs per chromosome and individual

```{r, echo=F}
roh$Chromosome <- gsub('chr','',roh$Chromosome)
roh$Chromosome <- as.numeric(roh$Chromosome)

roh$Chromosome <- factor(roh$Chromosome)
roh$Chromosome <- ordered(roh$Chromosome, levels = 1:22)

roh %>% group_by(sample_id,Chromosome,Range) %>%
  arrange(desc(Chromosome)) %>%
  summarise(SROH_Mb = sum(Length)/1e+06) %>%
  ggplot(aes(x=Chromosome,y=SROH_Mb,col=Range)) +
  geom_jitter() +
  theme_classic() +
  ylab('SROH (Mb)') +
  xlab('Chromosome') +
  scale_color_manual(values=group.colors)

kable(summary(
  roh %>% group_by(sample_id,Chromosome) %>%
  summarise(mean_SROH_Mb = sum(Length)/1e+06) %>%
    select(mean_SROH_Mb)
  )
  
)
```


3. Sum of ROH length per range

```{r SROH, echo=FALSE}
sroh <- roh %>% filter(Chromosome != 'chrX' & Chromosome != 'chrY') %>%
  filter(Length >= 2e+06 & Length <= 5e+06 | Length >= 10e+06 & Number_of_markers >= 50)
  
sroh_sum <- sroh %>% group_by(Range,sample_id) %>% summarise(SROH = sum(Length)/1e+06)

sroh_sum %>% ggplot(aes(y=Range,x=SROH)) +
 geom_violin(fill='#48C095', col='#27384A') +
 geom_boxplot(width=0.1) +
 theme_classic() +
 ylab('Range') +
 xlab('SROH (Mb)')

kable(
  summary(sroh_sum %>% 
            pivot_wider(names_from = Range, values_from = SROH) %>%
            select(-sample_id) %>% na.omit()
          )
)


```


4. Relationship between number of ROHs and total length of genome covered by them

```{r cummulative, echo=FALSE} 
croh <- roh %>% filter(Chromosome != 'chrX' & Chromosome != 'chrY' & Number_of_markers >= 50) 

croh_sum <- croh %>% group_by(sample_id) %>% 
  summarise(n = n(), sum = sum(Length)/1e+06)

croh_sum %>% ggplot(aes(y=n,x=sum)) +
  geom_point(col='#48C095') +
  theme_classic() +
  ylab('Number of ROH') +
  xlab('Total Length in ROH (Mb)')
```


5. ROH genome coverage 

```{r genome_coverage, echo=FALSE}
chr <- 21
coverage <- roh %>% filter(Chromosome %in% chr) %>%
  arrange(Chromosome,Start)
coverage$Chromosome <- paste(coverage$Chromosome,sep='')
coverage$x_from <- paste(coverage$Chromosome,coverage$Start,sep='')
coverage$x_end <- paste(coverage$Chromosome,coverage$End,sep='')



coverage %>% ggplot() +
  geom_segment(aes(x = x_from, y = sample_id, xend = x_end, 
                                 yend = sample_id), col='#48C095') +
  theme_classic() +
  theme(axis.text.x = element_blank(),axis.text.y = element_blank()) +
  xlab('Position on chromosome 21') +
  ylab('Individual')


```

```{r genome_coverage_heatmap, echo=FALSE}
library(GenomicRanges)
c = coverage(GRanges(seqnames = roh$Chromosome, 
                     ranges=IRanges(roh$Start, roh$End)))

window.size=1e6
m = matrix(nrow = 2*length(names(c)), 
           ncol=sum(runLength(c[[1]]))/window.size)

for (chr in 1:22) { #seqnames(gr)) {
  max = sum(runLength(c[[chr]]))
  sampling.points = IRanges(start=c(1:(max/1e6))*1e6, width = 1)  
  a = aggregate(c[[chr]], sampling.points, mean)
  m[(chr*2)-1,1:length(a)] = a
}

image(t(m[dim(m)[1]:1,]), xaxt='n', yaxt='n')

```



6. % of ROhs per category in sample

```{r roh_count, echo=FALSE}
suppressMessages(
roh_count <- roh  %>% 
 group_by(Range,sample_id,.drop = F) %>% 
 summarise(n=n()) %>% arrange(Range,n)
)

roh_count %>% 
 ggplot(aes(x=reorder(sample_id,-n),y=n,fill=Range)) +
 geom_bar(position="fill", stat="identity",width = 1) +
 scale_fill_manual(values=group.colors) +
 theme_classic() +
 xlab('Individuals') +
 ylab('% of ROHs count') +
 theme(axis.text.x = element_blank())
```

7. ROH length sum

```{r roh_sum, echo=FALSE}
suppressMessages(
roh_sum <- roh  %>% 
 group_by(Range,sample_id,.drop = F) %>% 
 summarise(sum=sum(Length)) %>% arrange(Range,sum)
)

roh_sum %>% 
 ggplot(aes(x=reorder(sample_id,-sum),y=sum,fill=Range)) +
 geom_bar(position="fill", stat="identity",width = 1) +
 scale_fill_manual(values=group.colors) +
 theme_classic() +
 xlab('Individuals') +
 ylab('% of ROHs length sum') +
 theme(axis.text.x = element_blank())
```

<!-- 7. Number of ROHs per sample -->

<!-- ```{r roh_count_per_sample, echo=FALSE}  -->
<!-- suppressMessages( -->
<!-- roh_avg <- roh %>% group_by(sample_id, Range) %>% -->
<!--   summarise(avg_length=mean(Length), n=n()) %>% -->
<!--   arrange(Range,n) -->
<!-- ) -->

<!-- roh_avg %>% -->
<!--   ggplot(aes(x=1:nrow(roh_avg),y=n)) + -->
<!--   geom_point(col='#48C095',alpha=0.5) + -->
<!--   facet_wrap(~Range, nrow = 3,scales = 'free') +  -->
<!--   ylab('Number of ROHs') + -->
<!--   xlab('Individual') + -->
<!--   theme_classic() + -->
<!--   theme(axis.text.x = element_blank()) -->

<!-- ``` -->

<!-- 8. Average ROHs per sample -->

<!-- ```{r average_roh_per_sample, echo=FALSE}  -->
<!-- roh_avg <- roh %>% group_by(sample_id, Range) %>% -->
<!--   summarise(avg_length=mean(Length), n=n()) %>% -->
<!--   arrange(Range,avg_length) -->

<!-- roh_avg %>% -->
<!--   ggplot(aes(x=1:nrow(roh_avg),y=avg_length)) + -->
<!--   geom_point(col='#48C095',alpha=0.5) + -->
<!--   facet_wrap(~Range, nrow = 3,scales = 'free') +  -->
<!--   ylab('Mean ROH length') + -->
<!--   xlab('Individual') + -->
<!--   theme_classic() + -->
<!--   theme(axis.text.x = element_blank()) -->

<!-- ``` -->

8. Cummulative sum

```{r echo=FALSE}
roh_cummulative <- roh %>% arrange(Range) %>% group_by(sample_id) %>%
  summarise(cummulative_length = cumsum(Length))

roh_cummulative %>%
  ggplot(aes(x=sample_id,y=cummulative_length)) +
  geom_point()

```

9. Cosanguinity in population

```{r echo=FALSE}
froh <- read.table('../output/roh/froh.output',header=T,sep='\t')

froh %>% ggplot(aes(x=1,y=Froh_genome)) +
    geom_violin(fill='#48C095', col='#27384A') +
  geom_boxplot(width=0.1) +
     theme_classic() + theme(axis.text.x = element_blank()) +
  xlab('')

```